---
title: "Sense of Belonging to Local Community Among Canadian Youths"
author: "Lawson Hung"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Install the Survey Package 
```{r, include = TRUE, message = FALSE, warning = FALSE}

install.packages(c("survey"))
```

### Load the Required Libraries
```{r, include = TRUE, message = FALSE, warning = FALSE}

library(mosaic)
library(tidyverse)
library(survey)
library(knitr)

```

### Import the Data File into R
```{r import data, include = TRUE, message = FALSE, warning = FALSE}

CCHS <- read.csv("CCHS.csv")
attach(CCHS)

```

### Convert Character Variables to Numeric Variables
```{r Convert Character to Numeric, include = TRUE, message = FALSE, warning = FALSE}

CCHS.Numeric <- CCHS %>% mutate_if(is.character, as.numeric)

```

### Filter Data by Youths Population
```{r Filter, include = TRUE, message = FALSE, warning = FALSE}

## Filtering by age Range (2 = 15-17, 3 = 18-19, 4 = 20-24, 5 = 25-29)

CCHS.Filtered <- CCHS.Numeric %>% filter(dhhgage > 1 & dhhgage <= 5)

```

### Select Variables and Drop NAs
```{r Select Variables and Drop NAs, include = TRUE, message = FALSE, warning = FALSE}

CCHS.Subset <- CCHS.Filtered %>% 
  select(GEN_030, DHH_SEX, dhhgage) %>% 
  drop_na(GEN_030, DHH_SEX, dhhgage)

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Re-level Data Values
```{r Re-level Values, include = TRUE, message = FALSE, warning = FALSE}

# Sense of Belonging to Local Community: 
## Levels: Strong, Weak
# Gender:
## Levels: Female, Male
# Age: 
## Levels: 15-17, 18-19, 20-24, 25-29
CCHS.Subset <- CCHS.Subset %>% 
  mutate(Belonging.Community = case_when(GEN_030 <=2 ~ "Strong",
                                         GEN_030 >=3 ~ "Weak"), 
        Gender = case_when(DHH_SEX == 1 ~ "Male", 
                           DHH_SEX == 2 ~ "Female"),
        Age = case_when(dhhgage == 2 ~ "15-17",
                         dhhgage == 3 ~ "18-19",
                         dhhgage == 4 ~ "20-24",
                         dhhgage == 5 ~ "25-29"))                            
attach(CCHS.Subset)

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Make a Pseudo-population
```{r Pseudo-population, include = TRUE, message = FALSE, warning = FALSE}

# Input your seed number below for reproducibility.
set.seed(797)

# Take a sample of 5000 rows without replacement.
# Assume this is population distribution.
CCHS.pop <- CCHS.Subset[sample(1:nrow(CCHS.Subset), size = 1000, replace = FALSE), ]

# Create a data frame to store the sampling results in. 
CCHS.pop <- as.data.frame(CCHS.pop)

# Attach the data frame. 
attach(CCHS.pop)

# Detach the original data frame.
detach(CCHS.Subset)
```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Proportional Allocation
```{r Proportional Allocation, include = TRUE, message = FALSE, warning = FALSE}

# With proportional allocation, the stratum sample sizes are proportional to 
# the population stratum sizes Ni. 

# Multiply Ni/N by the sample size.
 age.pop <- table(Age); age.pop

# Sample size is 100
n = 100 

# Population size is 1000
N <- nrow(CCHS.pop); N

# Obtain proportional allocation
prop.alloc <- n*age.pop/N
prop.alloc

# Round to nearest integer
prop.alloc_int <- round(prop.alloc)
prop.alloc_int

# Check that stratum sample sizes sum to 100
sum(prop.alloc_int)

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Stratified Random Sample: Using the slice_sample() function from the dplyr package 
```{r Stratified Sample, include = TRUE, message = FALSE, warning = FALSE}

# Set seed for reproducibility
set.seed(797)

# Define the number of samples per age
# Create a vector with the number of samples to draw from each stratum
# You need to change stratum sizes based on the result of your seed number
sample_sizes <- c("15-17" = 20, "18-19" = 11, "20-24" = 29, "25-29" = 40)


# Perform stratified sampling
# Starts a pipe (%>%) that takes CCHS.pop as input and applies a series of operations
strsample <- CCHS.pop %>%
  # Add original row numbers
  mutate(row_id = row_number()) %>%
  # Groups the dataframe by the stratifying variable Age
  group_by(Age) %>%
  # group_modify applies a function to each group:
  # .x is the current group’s data (all rows of that Age)
  # .y contains the group key (here Age).
  # slice_sample(.x, n = sample_sizes[.y$sex[1]]) randomly selects exactly 
  # the correct number of rows from that group according to sample_sizes
  # slice_sample() automatically samples without replacement by default (replace = FALSE)
  group_modify(~ slice_sample(.x, n = sample_sizes[.y$Age[1]])) %>%
  # Removes the grouping, returning a regular dataframe instead of a grouped one
  ungroup()

attach(strsample)

# Check the sample sizes
table(strsample$Age)

# Keep the following variables
strsample <- strsample %>% select(row_id, Age, Belonging.Community)

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Distribution of Sense of Beloning to Local Community
```{r Sense of Belonging Distn, include = TRUE, message = FALSE, warning = FALSE}

kable(tally(~ Belonging.Community, margin = TRUE, data = strsample))

strsample %>%
  # count each category
  count(Belonging.Community) %>%                          
  mutate(
    # calculate percentage
    Percent = n / sum(n) * 100,      
    # combine count and percent
    Label = paste0(n, " (", round(Percent, 2), "%)") 
  ) %>%
  select(Belonging.Community, Label) %>%
  kable(
    col.names = c("Sense", "Count (Percent)"),
    caption = "Community Belonging"
  )

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Perception of Sense of Belonging to Local Community by Age
```{r Sense of Belonging by Age, include = TRUE, message = FALSE, warning = FALSE}

kable(tally(~ Belonging.Community + Age, margin = TRUE, data = strsample))

strsample %>%
  # count each combination
  count(Age, Belonging.Community) %>%              
  group_by(Age) %>%
  # calculate row percentages
  mutate(Percent = n / sum(n) * 100) %>%     
  ungroup() %>%
  # combine count + percent
  mutate(Result = paste0(n, " (", round(Percent, 2), "%)")) %>%  
  select(Age, Belonging.Community, Result) %>%
  tidyr::pivot_wider(names_from = Belonging.Community,
                     values_from = Result)  %>%
  knitr::kable()

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Sense of Belonging to Local Community by Age
```{r Bar Plot of Sense of Belonging to Local Community by Age, include = TRUE, message = FALSE, warning = FALSE}

strsample %>%
  # count each combination
  count(Age, Belonging.Community) %>%                  
  group_by(Age) %>%
  # compute conditional percentages
  mutate(percent = n / sum(n) * 100) %>%         
  ggplot(aes(x = factor(Age), 
             y = percent, 
             fill = factor(Belonging.Community))) +
  # side-by-side bars
  geom_col(position = "dodge") +                
  scale_y_continuous(labels = scales::percent_format(scale = 1)) + 
  # custom colors
  scale_fill_manual(
    values = c("Strong" = "seagreen3", "Weak" = "mediumpurple3")  
  ) +
  labs(
    x = "Age Group",
    y = "Percent within Age Group",
    fill = "Sense of Belonging",
    title = "Sense of Belonging to Local Community by Age"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    # light green background
    plot.background = element_rect(fill = "honeydew2", color = NA),  
    # same for panel
    panel.background = element_rect(fill = "honeydew2", color = NA), 
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_blank()
  )

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Create A Binary Variable for Sense of Belonging to Local Community 
```{r Create a Binary Variable, include = TRUE, message = FALSE, warning = FALSE}

# Create a variable Sense.Belonging, 
# replicate 0 n times (number of rows in the sample) (n = 100)
strsample$Sense.Belonging <- rep(0, nrow(strsample))
# If Sense.Community in strsample == Strong, assign value 1, otherwise (if Weak), remain 0
strsample$Sense.Belonging[strsample$Belonging.Community == "Strong"] <- 1

```

### Specify the Survey Design
```{r Stratified Design, include = TRUE, message = FALSE, warning = FALSE}

# Create a variable containing population stratum sizes, for use in fpc
strsample <- strsample %>%
  mutate(
    # Change the population sizes according to the result from your seed number
    popsize = case_when(
      Age == "15-17" ~ 224,
      Age == "18-19" ~ 109,
      Age == "20-24" ~ 270,
      Age == "25-29" ~ 397
    ))


# Check the result
strsample %>%
  count(Age, popsize)

# Input design information for strsample
# Use svydesign() function in survey package
# Use ~1 for id argument to indicate no clusters
# strata argument gives the variable name(s) containing the stratification information
# here, the stratification variable is age
# fcp argument specifies the variable that contains information for calculating 
# the finite population correction (fcp) in each stratum.
stratified.design <- svydesign(id = ~1, strata = ~Age, fpc = ~popsize, 
                               data = strsample)

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Calculate Proportion, SE and Confidence Interval
```{r Estimate Population Proportion, include = TRUE, message = FALSE, warning = FALSE}

est.prop <- svymean(~Sense.Belonging, stratified.design)
est.prop
confint(est.prop)

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Calculate Total, SE and Confidence Interval
```{r Estimate Population Total, include = TRUE, message = FALSE, warning = FALSE}

est.total <- svytotal(~Sense.Belonging, stratified.design)
est.total
confint(est.total)

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Alternative Way to Calculate Estimated Proportion and Total
```{r Alterantive Way, include = TRUE, message = FALSE, warning = FALSE}

# Create a factor variable sense.belonging
strsample$sense.belonging <- factor(strsample$Sense.Belonging)

# Return the svydesign because the data set has a new variable
stratified.design <- svydesign(id = ~1, strata = ~Gender, fpc = ~popsize, 
                               data = strsample)

# Calculate Proportion, SE and Confidence Interval
est.prop <- svymean(~sense.belonging, stratified.design)
est.prop
confint(est.prop)

# Calculate Total, SE and Confidence Interval
est.total <- svytotal(~sense.belonging, stratified.design)
est.total
confint(est.total)

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

### Calculate Stratum Statistics
```{r Stratum Statistics, include = TRUE, message = FALSE, warning = FALSE}

est.prop.age <- svyby(~sense.belonging, by = ~Age, stratified.design, svymean, keep.var = TRUE)
est.prop.age
confint(est.prop.age)

```